---
title: "Respiratory Risk & Air Pollution"
subtitle: "State-Level Analysis — Reproducible Notebook"
author: "Adham Abdelhalim · Odinaka Obioha · Chinmoy Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    number_sections: true
    df_print: paged
    theme: flatly
    highlight: tango
    code_folding: hide
    includes:
      in_header: report_header.html
editor_options: null
chunk_output_type: inline
---

### 1.1 Environment Setup
First, we load the core libraries required for data wrangling and advanced visualizations.

```{r setup, message=FALSE, warning=FALSE}
# Load core libraries
library(tidyverse)  # Includes dplyr, purrr, tidyr, stringr, readr, ggplot2
library(ggtext)     # For advanced text formatting in plots
library(viridis)    # For colorblind-friendly palettes
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 150,
  fig.retina = 2,
  fig.width = 9,
  fig.height = 5.5,
  out.width = "100%"
)

print("Libraries loaded successfully.")
```
### 1.2 File Targeting & Standardization
Here, we scan the directory for our specific CSV files and create a helper function to ensure state names match perfectly across all datasets.

```{r file_setup}
# 1. Target the files using pattern matching (regex)
mort_files     <- list.files(pattern = "Respiratory Mortality.*\\.csv$")
age_std_files  <- list.files(pattern = "Age standardised mortality.*\\.csv$")
no2_files      <- list.files(pattern = "Nitrogen dioxide.*\\.csv$")
pm10_files     <- list.files(pattern = "Particulate matter.*\\.csv$")

# 2. Create a helper function to fix spelling inconsistencies
standardize_states <- function(data) {
  data %>%
    mutate(State = case_when(
      str_detect(State, "Baden-W.rttemberg") ~ "Baden-Wurttemberg",
      State == "Mecklenburg-West Pomerania" ~ "Mecklenburg Western Pomerania",
      State == "Saxony-Anhalt" ~ "Saxony Anhalt",
      TRUE ~ State
    ))
}

# 3. Verify files found
cat("Found", length(mort_files), "Mortality files\n")
cat("Found", length(age_std_files), "Age-standardised mortality files\n")
cat("Found", length(no2_files), "NO2 files\n")
cat("Found", length(pm10_files), "PM10 files\n")
```

### 1.3 Process Mortality Data
Here, we extract the year, pivot the states, fix data types, and split the data: saving the national total separately while filtering out non-state regional aggregates.

```{r process_mortality, warning=FALSE}
# 1. First, load everything into a temporary dataframe
df_mortality_all <- map_dfr(mort_files, function(file) {
  
  yr <- as.numeric(str_extract(file, "\\d{4}"))
  raw_df <- read_delim(file, delim = ";", skip = 9, show_col_types = FALSE)
  condition_col <- names(raw_df)[1]
  
  raw_df %>%
    filter(str_detect(!!sym(condition_col), "J00-J99")) %>%
    pivot_longer(
      cols = -1, 
      names_to = "State", 
      values_to = "Mortality_Rate",
      values_transform = list(Mortality_Rate = as.character)
    ) %>%
    mutate(
      Year = yr,
      Mortality_Rate = as.numeric(str_replace_all(Mortality_Rate, ",", ""))
    ) %>%
    select(State, Year, Mortality_Rate)
}) %>% 
  standardize_states()

# 2. Save the Germany national total separately for later use
df_germany_total <- df_mortality_all %>%
  filter(State == "Germany")

# 3. Create the final clean dataset by dropping the junk rows and the Germany total
df_mortality <- df_mortality_all %>%
  filter(!State %in% c("Germany", "Former federal territory and East Berlin", 
                       "New federal states excluding East Berlin", "...21"))

# Verify the clean 16-state dataset
print(df_mortality, n = 6)
```

### 1.3b Process Age-Standardised Mortality
Official GBE age-standardised rates (European standard population 2013). Excludes Germany and the two regional aggregate rows.

```{r process_age_std, warning=FALSE}
df_age_std <- map_dfr(age_std_files, function(file) {
  yr <- as.numeric(str_extract(file, "\\d{4}"))
  suppressMessages(read_delim(
    file, delim = ";", skip = 12, n_max = 18,
    col_names = c("State", "Deaths_asd", "Deaths_per_100k", "Mort_age_std"),
    col_types = cols(.default = col_character()),
    locale = locale(encoding = "latin1")
  )) %>%
    mutate(
      Year = yr,
      State = str_trim(State),
      Mort_age_std = as.numeric(str_replace(Mort_age_std, ",", "."))
    ) %>%
    filter(!State %in% c(
      "Germany",
      "Former federal territory and East Berlin",
      "New federal states excluding East Berlin"
    )) %>%
    select(State, Year, Mort_age_std)
}) %>%
  standardize_states()

print(df_age_std, n = 6)
```
### 1.4 Process Pollution Data & Merge
Now we handle the NO2 and PM10 files using a unified helper function, and join everything into one final, tidy master dataset.

```{r process_pollution_and_merge, warning=FALSE}
# 1. Helper Function: Clean Pollution Data
process_pollution <- function(files, value_name) {
  map_dfr(files, function(file) {
    yr <- as.numeric(str_extract(file, "\\d{4}"))
    
    # Read CSV, forcing column types to avoid parsing errors
    raw_df <- read_delim(file, delim = ";", show_col_types = FALSE)
    
    # Identifying column names dynamically
    state_col <- names(raw_df)[str_detect(names(raw_df), "(?i)State|network")]
    val_col <- names(raw_df)[str_detect(names(raw_df), "(?i)Annual mean value")]
    
    raw_df %>%
      rename(State = all_of(state_col), Mean_Value = all_of(val_col)) %>%
      # Filter out 'UBA' and rows containing asterisks
      filter(
        !str_detect(State, "UBA"),
        !str_detect(State, "\\*")
      ) %>%
      # Ensure numeric values (handling comma decimals)
      mutate(Mean_Value = as.numeric(str_replace(Mean_Value, ",", "."))) %>%
      group_by(State) %>%
      summarize(!!value_name := mean(Mean_Value, na.rm = TRUE), .groups = "drop") %>%
      mutate(Year = yr)
  }) %>% standardize_states()
}

# 2. Process NO2 and PM10 files
df_no2  <- process_pollution(no2_files, "NO2_Level")
df_pm10 <- process_pollution(pm10_files, "PM10_Level")

# 3. Merge into Tidy Master Dataframe
master_df <- df_mortality %>%
  left_join(df_age_std, by = c("State", "Year")) %>%
  left_join(df_no2, by = c("State", "Year")) %>%
  left_join(df_pm10, by = c("State", "Year")) %>%
  drop_na()

# Save for 2.Rmd interactive notebook
saveRDS(master_df, "master_df.rds")

# Display the first 6 rows of the final merged dataset
head(master_df)
```
### 2.1 Plot 1: Longitudinal Trend Analysis
Visualizing the 5-year respiratory mortality trend, isolating Bavaria to demonstrate localized severity.

```{r plot_1_longitudinal, fig.width=9, fig.height=6}
# Define our consultant-style theme
theme_consultant <- theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    # ggtext allows us to use HTML/Markdown in the subtitle!
    plot.subtitle = element_markdown(size = 12, color = "grey30", margin = margin(b = 10)),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# Generate Plot 1
plot1 <- master_df %>%
  mutate(
    # Create highlighting logic for Bavaria
    Highlight = ifelse(State == "Bavaria", "Bavaria", "Other"),
    LineSize = ifelse(State == "Bavaria", 1.2, 0.6),
    Alpha = ifelse(State == "Bavaria", 1, 0.4)
  ) %>%
  ggplot(aes(x = Year, y = Mort_age_std, group = State)) +
  geom_line(aes(color = Highlight, size = LineSize, alpha = Alpha)) +
  # Use a distinct, colorblind-friendly orange for Bavaria
  scale_color_manual(values = c("Bavaria" = "#D55E00", "Other" = "grey70")) +
  scale_size_identity() +
  scale_alpha_identity() +
  scale_x_continuous(breaks = 2019:2023) +
  labs(
    title = "Age-Standardised Respiratory Mortality Trends in Germany (2019-2023)",
    subtitle = "Highlighting the <span style='color:#D55E00;'>**Bavarian**</span> trajectory against all other 15 federal states.",
    x = "Year",
    y = "Age-standardised deaths per 100,000"
  ) +
  theme_consultant

# Display the plot
plot1
```
### 2.2 Plot 2: Year-over-Year Correlation
Here we use a faceted scatterplot to prove that the correlation between environmental pollutants and respiratory mortality remains consistent regardless of the year.

```{r plot_2_scatter, warning=FALSE, message=FALSE}
# Generate Plot 2
plot2 <- master_df %>%
  ggplot(aes(x = NO2_Level, y = Mort_age_std)) +
  # Add scatter points colored by PM10
  geom_point(aes(color = PM10_Level), alpha = 0.8, size = 2.5) +
  # Add the linear regression trend line
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", size = 0.8) +
  # Apply the colorblind-friendly continuous palette
  scale_color_viridis_c(option = "viridis", name = "PM10 Level (µg/m³)") +
  # Split into 5 side-by-side plots based on the Year
  facet_wrap(~Year, nrow = 1) +
  labs(
    title = "Impact of NO2 and PM10 on Respiratory Mortality",
    subtitle = "Consistent linear correlation observed year-over-year.",
    x = "Average NO2 Exposure (µg/m³)",
    y = "Age-standardised deaths per 100,000"
  ) +
  theme_consultant +
  theme(
    strip.text = element_text(face = "bold", size = 12), # Make the year labels stand out
    legend.key.width = unit(2, "cm")                     # Widen the legend for readability
  )

# Display the plot
plot2
```
### 3.1 Distribution of Pollutants Over Time
```{r plot_boxplot, fig.width=8, fig.height=5, warning=FALSE}
# Pivot the pollutants into a longer format for side-by-side plotting
plot_dist <- master_df %>%
  select(State, Year, NO2_Level, PM10_Level) %>%
  pivot_longer(cols = c(NO2_Level, PM10_Level), names_to = "Pollutant", values_to = "Concentration") %>%
  ggplot(aes(x = factor(Year), y = Concentration, fill = Pollutant)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 2) +
  scale_fill_manual(
    values = c("NO2_Level" = "#E69F00", "PM10_Level" = "#56B4E9"),
    labels = c("NO2", "PM10")
  ) +
  labs(
    title = "Variance in State-Level Pollutant Concentrations (2019-2023)",
    x = "Year",
    y = "Concentration (µg/m³)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", legend.title = element_blank())

plot_dist
```
### 3.2 Magnitude of Change: 2019 vs 2023
```{r plot_dumbbell, fig.width=9, fig.height=7, warning=FALSE}
plot_dumbbell <- master_df %>%
  # Filter for only start and end years
  filter(Year %in% c(2019, 2023)) %>%
  select(State, Year, Mort_age_std) %>%
  # Pivot wider to get 2019 and 2023 into their own columns
  pivot_wider(names_from = Year, values_from = Mort_age_std, names_prefix = "Yr_") %>%
  # Reorder states by their 2019 baseline for a cleaner visual hierarchy
  ggplot(aes(y = reorder(State, Yr_2019))) +
  # Draw the connecting line
  geom_segment(aes(x = Yr_2019, xend = Yr_2023, yend = State), color = "grey70", linewidth = 1) +
  # Add the start and end points
  geom_point(aes(x = Yr_2019, color = "2019"), size = 3) +
  geom_point(aes(x = Yr_2023, color = "2023"), size = 3) +
  scale_color_manual(values = c("2019" = "#0072B2", "2023" = "#D55E00")) +
  labs(
    title = "Shift in Age-Standardised Respiratory Deaths (2019 vs. 2023)",
    x = "Deaths per 100,000 (age-standardised)",
    y = NULL,
    color = "Year"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    legend.position = "top"
  )

plot_dumbbell
```