---
title: "Interactive Plots"
subtitle: "Linked Visualisations — Reproducible Notebook"
author: "Adham Abdelhalim · Odinaka Obioha · Chinmoy Bora"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: hide
    includes:
      in_header: report_header.html
---
    
```{r load-data, include=FALSE}

# This tells the knitter to grab the file you just saved
master_df <- readRDS("master_df.rds")
```

# Interactive Risk Mapping

```{r setup, include=FALSE}
# Global settings for crisp rendering
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  dpi = 150,
  fig.retina = 2,
  fig.width = 9,
  fig.height = 5.5,
  out.width = "100%"
)

# Load our interactive arsenal
library(tidyverse)
library(plotly)          # For interactive charts 
library(leaflet)         # For interactive maps
library(sf)              # For spatial data (polygons/borders)
library(rnaturalearth)   # For downloading German map boundaries
library(scales)
```



### 1.1 Geographic Data Preparation

Here we fetch the geospatial boundaries for Germany and merge them with our 2023 environmental dataset.

```{r prep_map_data}
# 1. Download the geographic borders of German states
germany_borders <- ne_states(country = "germany", returnclass = "sf")

# 2. Create a translation dictionary to match our master_df
germany_borders <- germany_borders %>%
  mutate(State = case_when(
    name == "Bayern" ~ "Bavaria",
    name == "Baden-Württemberg" ~ "Baden-Wurttemberg",
    name == "Nordrhein-Westfalen" ~ "North Rhine-Westphalia",
    name == "Hessen" ~ "Hesse",
    name == "Sachsen" ~ "Saxony",
    name == "Niedersachsen" ~ "Lower Saxony",
    name == "Rheinland-Pfalz" ~ "Rhineland-Palatinate",
    name == "Thüringen" ~ "Thuringia",
    name == "Sachsen-Anhalt" ~ "Saxony Anhalt",
    name == "Mecklenburg-Vorpommern" ~ "Mecklenburg Western Pomerania",
    TRUE ~ name # Berlin, Bremen, Hamburg, Saarland, Schleswig-Holstein, Brandenburg match perfectly
  ))

# 3. Filter our master_df for 2023 and merge it onto the map shapes
# (Note: Assuming 'master_df' is loaded in your environment from the previous session)
map_data_2023 <- germany_borders %>%
  left_join(filter(master_df, Year == 2023), by = "State")

print("Map geometries and environmental data successfully merged!")
```

### 1.2 Interactive Environmental Risk Map (2023)

```{r leaflet_map, fig.height=2, fig.width=9, out.width="100%", warning=FALSE, message=FALSE}
# 1. Define a color palette (Yellow to Dark Red based on age-standardised mortality)
color_pal <- colorNumeric(palette = "YlOrRd", domain = map_data_2023$Mort_age_std)

# 2. Design the HTML hover tooltip
hover_labels <- sprintf(
  "<strong>%s</strong><br/>
  Age-std deaths/100k: %.1f<br/>
  PM10 Level: %g µg/m³<br/>
  NO2 Level: %g µg/m³",
  map_data_2023$State, 
  map_data_2023$Mort_age_std, 
  round(map_data_2023$PM10_Level, 1),
  round(map_data_2023$NO2_Level, 1)
) %>% lapply(htmltools::HTML)

# 3. Build the Leaflet Map
interactive_map <- leaflet(data = map_data_2023) %>%
  # Add a clean, minimal base map (like Google Maps, but better for data)
  addProviderTiles(providers$CartoDB.Positron) %>%
  # Add the state polygons
  addPolygons(
    fillColor = ~color_pal(Mort_age_std),
    weight = 1.5,
    opacity = 1,
    color = "white", # White borders between states
    fillOpacity = 0.8,
    highlightOptions = highlightOptions(
      weight = 4, 
      color = "#666", 
      fillOpacity = 1, 
      bringToFront = TRUE
    ),
    label = hover_labels,
    labelOptions = labelOptions(
      style = list("font-weight" = "normal", padding = "8px 12px"),
      textsize = "14px",
      direction = "auto"
    )
  ) %>%
  # Add the legend in the bottom right
  addLegend(
    pal = color_pal, 
    values = ~Mort_age_std, 
    opacity = 0.8, 
    title = "2023 Age-std deaths/100k",
    position = "bottomright"
  )

# Render!
interactive_map
```

### Normalizing data
```{r finalize-data-patch, message=FALSE, warning=FALSE}
# master_df from 1.Rmd already includes official GBE age-standardised rates
master_df <- readRDS("master_df.rds")

head(master_df %>% select(State, Year, Mortality_Rate, Mort_age_std), 3)
```


###1.3 Interactive Correlation Scatter Plot
```{r correlation-plot-interactive, message=FALSE, warning=FALSE, out.width="100%", fig.height=5}
library(ggplot2)
library(plotly)

# 1. Transform the dataframe into a reactive object grouped by State
linked_data <- highlight_key(master_df, ~State)

# 2. PM10 Plot (Using the linked_data)
gg_pm10 <- ggplot(linked_data, aes(x = PM10_Level, y = Mort_age_std, color = State)) +
  geom_point(aes(text = sprintf("<b>%s (%d)</b><br>PM10: %.1f µg/m³<br>Age-std/100k: %.1f",
                                State, Year, PM10_Level, Mort_age_std)),
             size = 3, alpha = 0.8) +
  # Note: We use the raw master_df for the trendline so it doesn't try to calculate a line for just the highlighted points
  geom_smooth(data = master_df, aes(group = 1), method = "lm", color = "#444444", linetype = "dashed", linewidth = 1, se = FALSE) +
  labs(x = "PM10 Concentration (µg/m³)", y = "Age-standardised deaths per 100,000") +
  theme_bw() + 
  theme(legend.position = "none", panel.grid.minor = element_blank())

# 3. NO2 Plot (Using the linked_data)
gg_no2 <- ggplot(linked_data, aes(x = NO2_Level, y = Mort_age_std, color = State)) +
  geom_point(aes(text = sprintf("<b>%s (%d)</b><br>NO2: %.1f µg/m³<br>Age-std/100k: %.1f",
                                State, Year, NO2_Level, Mort_age_std)),
             size = 3, alpha = 0.8) +
  geom_smooth(data = master_df, aes(group = 1), method = "lm", color = "#444444", linetype = "dashed", linewidth = 1, se = FALSE) +
  labs(x = "NO2 Concentration (µg/m³)", y = "") + 
  theme_bw() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

# 4. Combine and enforce the highlight behavior
subplot(ggplotly(gg_pm10, tooltip = "text"), 
        ggplotly(gg_no2, tooltip = "text"), 
        nrows = 1, titleX = TRUE, titleY = TRUE, margin = 0.08) %>%
  layout(
    title = list(text = "Pollutant Exposure vs. Age-Standardised Respiratory Mortality (Hover to group by state)",
                 font = list(size = 14)),
    showlegend = FALSE,
    height = 480,
    margin = list(l = 80, r = 30, t = 60, b = 20),
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  ) %>%
  highlight(
    on = "plotly_hover", 
    off = "plotly_doubleclick", 
    opacityDim = 0.1,
    hoverinfo = "text"
  ) %>%
  config(displayModeBar = FALSE)
```
### Longitudinal trends

```{r longitudinal-trends, message=FALSE, warning=FALSE, out.width="100%", fig.height=5}
library(ggplot2)
library(plotly)

# 1. Establish the linked data structure mapped by State
linked_time_data <- highlight_key(master_df, ~State)

# 2. PM10 Longitudinal Trajectory Plot
gg_ts_pm10 <- ggplot(linked_time_data, aes(x = Year, y = PM10_Level, color = State, group = State)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(aes(text = sprintf("<b>%s</b><br>Year: %d<br>PM10: %.1f µg/m³", 
                                State, Year, PM10_Level)),
             size = 2) +
  scale_x_continuous(breaks = 2019:2023) + # Ensures crisp integer year ticks
  labs(x = "Year", y = "PM10 Concentration (µg/m³)") +
  theme_bw() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

# 3. NO2 Longitudinal Trajectory Plot
gg_ts_no2 <- ggplot(linked_time_data, aes(x = Year, y = NO2_Level, color = State, group = State)) +
  geom_line(linewidth = 1, alpha = 0.8) +
  geom_point(aes(text = sprintf("<b>%s</b><br>Year: %d<br>NO2: %.1f µg/m³", 
                                State, Year, NO2_Level)),
             size = 2) +
  scale_x_continuous(breaks = 2019:2023) +
  labs(x = "Year", y = "NO2 Concentration (µg/m³)") +
  theme_bw() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

# 4. Combine into an interconnected subplot
subplot(ggplotly(gg_ts_pm10, tooltip = "text"), 
        ggplotly(gg_ts_no2, tooltip = "text"), 
        nrows = 1, titleX = TRUE, titleY = TRUE, margin = 0.08) %>%
  layout(
    title = list(text = "Longitudinal Air Pollution Trajectories across German States (2019–2023)",
                 font = list(size = 14)),
    showlegend = FALSE,
    height = 480,
    margin = list(l = 80, r = 30, t = 60, b = 20),
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  ) %>%
  highlight(
    on = "plotly_hover", 
    off = "plotly_doubleclick", 
    opacityDim = 0.1,               # Mutes non-targeted states to 10% opacity
    hoverinfo = "text"
  ) %>%
  config(displayModeBar = FALSE)
```
### The Dumbbell Plot: 2019 vs. 2023 Shift(PM10)

```{r dumbbell-plot, message=FALSE, warning=FALSE, out.width="100%", fig.height=6}

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

# 1. Reshape the data
dumbbell_data <- master_df %>%
  filter(Year %in% c(2019, 2023)) %>%
  select(State, Year, PM10_Level) %>%
  pivot_wider(names_from = Year, values_from = PM10_Level, names_prefix = "Y") %>%
  mutate(
    Change = Y2023 - Y2019,
    State = reorder(State, Change)
  )

# 2. Build the Directed Dumbbell Plot
gg_dumbbell <- ggplot(dumbbell_data) +
  # Draw the connecting baseline
  geom_segment(aes(x = Y2019, xend = Y2023, y = State, yend = State), 
               color = "gray80", linewidth = 1.5) +
               
  # THE UX FIX: Add a left-pointing arrow in the exact center of the bar
  geom_text(aes(x = (Y2019 + Y2023) / 2, y = State, label = "←"), 
            color = "gray40", size = 6, fontface = "bold") +
            
  # Draw the 2019 starting points (Orange)
  geom_point(aes(x = Y2019, y = State, 
                 text = sprintf("<b>%s</b><br>2019 PM10: %.1f µg/m³", State, Y2019)), 
             color = "#fc8d62", size = 4, alpha = 0.9) +
  # Draw the 2023 ending points (Blue)
  geom_point(aes(x = Y2023, y = State, 
                 text = sprintf("<b>%s</b><br>2023 PM10: %.1f µg/m³<br>Net Change: %+.1f µg/m³", State, Y2023, Change)), 
             color = "#8da0cb", size = 4, alpha = 0.9) +
  labs(x = "PM10 Concentration (µg/m³)", y = "", 
       title = "Magnitude of Shift in PM10 Levels (2019 to 2023)") +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank() 
  )

# 3. Convert to interactive widget
ggplotly(gg_dumbbell, tooltip = "text") %>%
  layout(
    height = 576,
    margin = list(l = 170, r = 30, t = 60, b = 20),
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  ) %>%
  config(displayModeBar = FALSE)
```


### The Dumbbell Plot: 2019 vs. 2023 Shift(NO2)
```{r dumbbell-plot-no2, message=FALSE, warning=FALSE, out.width="100%", fig.height=6}

library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)

# 1. Reshape the data for NO2 specifically
dumbbell_data_no2 <- master_df %>%
  filter(Year %in% c(2019, 2023)) %>%
  select(State, Year, NO2_Level) %>%
  pivot_wider(names_from = Year, values_from = NO2_Level, names_prefix = "Y") %>%
  mutate(
    Change = Y2023 - Y2019,
    State = reorder(State, Change) # Sorts the Y-axis by the magnitude of change for NO2
  )

# 2. Build the Directed Dumbbell Plot for NO2
gg_dumbbell_no2 <- ggplot(dumbbell_data_no2) +
  # Draw the connecting baseline
  geom_segment(aes(x = Y2019, xend = Y2023, y = State, yend = State), 
               color = "gray80", linewidth = 1.5) +
               
  # THE UX FIX: Add the left-pointing arrow in the exact center of the bar
  geom_text(aes(x = (Y2019 + Y2023) / 2, y = State, label = "←"), 
            color = "gray40", size = 6, fontface = "bold") +
            
  # Draw the 2019 starting points (Orange to maintain visual consistency)
  geom_point(aes(x = Y2019, y = State, 
                 text = sprintf("<b>%s</b><br>2019 NO2: %.1f µg/m³", State, Y2019)), 
             color = "#fc8d62", size = 4, alpha = 0.9) +
  # Draw the 2023 ending points (Blue to maintain visual consistency)
  geom_point(aes(x = Y2023, y = State, 
                 text = sprintf("<b>%s</b><br>2023 NO2: %.1f µg/m³<br>Net Change: %+.1f µg/m³", State, Y2023, Change)), 
             color = "#8da0cb", size = 4, alpha = 0.9) +
  labs(x = "NO2 Concentration (µg/m³)", y = "", 
       title = "Magnitude of Shift in NO2 Levels (2019 to 2023)") +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank() # Clean background
  )

# 3. Convert to interactive widget
ggplotly(gg_dumbbell_no2, tooltip = "text") %>%
  layout(
    height = 576,
    margin = list(l = 170, r = 30, t = 60, b = 20),
    plot_bgcolor  = "white",
    paper_bgcolor = "white"
  ) %>%
  config(displayModeBar = FALSE)
```


###Risk Matrix (Quadrant Bubble Chart)
```{r risk-matrix-2023-beautiful, message=FALSE, warning=FALSE, out.width="100%", fig.height=5.5}
library(dplyr)
library(ggplot2)
library(plotly)

# 1. Isolate 2023 and Calculate Medians
data_2023 <- master_df %>% filter(Year == 2023)
med_pm10 <- median(data_2023$PM10_Level, na.rm = TRUE)
med_no2  <- median(data_2023$NO2_Level, na.rm = TRUE)

# Calculate dynamic bounds to ensure the shading covers the whole top-right area
max_pm10 <- max(data_2023$PM10_Level) + 2
max_no2  <- max(data_2023$NO2_Level) + 2

# 2. Categorize the data by Quadrant
data_2023 <- data_2023 %>%
  mutate(Risk_Category = case_when(
    PM10_Level > med_pm10 & NO2_Level > med_no2 ~ "High Risk (Both)",
    PM10_Level <= med_pm10 & NO2_Level <= med_no2 ~ "Low Risk (Both)",
    TRUE ~ "Moderate Risk (Mixed)"
  ))

# 3. Build the Upgraded Bubble Chart
gg_bubble <- ggplot(data_2023, aes(x = PM10_Level, y = NO2_Level)) +
  
  # THE FIX 1: Draw the red danger zone natively in ggplot2 so it cannot be hidden
  annotate("rect", xmin = med_pm10, xmax = max_pm10, 
           ymin = med_no2, ymax = max_no2, 
           fill = "#e34a33", alpha = 0.08) +
  
  # The Crosshairs
  geom_vline(xintercept = med_pm10, linetype = "dashed", color = "#888888", linewidth = 0.8) +
  geom_hline(yintercept = med_no2, linetype = "dashed", color = "#888888", linewidth = 0.8) +
  
  # The Bubbles
  geom_point(aes(size = Mort_age_std, 
                 fill = Risk_Category, 
                 text = sprintf("<b>%s</b><br>Category: %s<br>PM10: %.1f<br>NO2: %.1f<br>Age-std/100k: %.1f", 
                                State, Risk_Category, PM10_Level, NO2_Level, Mort_age_std)),
             alpha = 0.85, shape = 21, color = "white", stroke = 1.2) +
  
  # Color palette
  scale_fill_manual(values = c("High Risk (Both)" = "#e34a33",     
                               "Moderate Risk (Mixed)" = "#fec44f", 
                               "Low Risk (Both)" = "#a1d99b")) +    
  
  # THE FIX 2: Explicitly name and format the size legend
  scale_size_continuous(range = c(4, 16), name = "Age-std/100k") + 
  
  # Hide the redundant color legend, but keep the size legend
  guides(fill = "none") +
  
  labs(
    title = "Co-Exposure Risk Matrix: PM10 vs. NO2 (2023)",
    x = paste0("PM10 Concentration (Median: ", round(med_pm10, 1), " µg/m³)"),
    y = paste0("NO2 Concentration (Median: ", round(med_no2, 1), " µg/m³)")
  ) +
  theme_bw() +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 14)
    # We removed legend.position="none" here so the size guide can render
  )

# 4. Convert to interactive widget
ggplotly(gg_bubble, tooltip = "text") %>%
  layout(
    height = 528,
    plot_bgcolor  = "white",
    paper_bgcolor = "white",
    margin = list(l = 80, r = 30, t = 60, b = 20)
  ) %>%
  add_annotations(
    x = max_pm10 - 0.5, y = max_no2 - 0.5,
    text = "Danger Zone", showarrow = FALSE,
    xanchor = "right", yanchor = "top", font = list(color = "#e34a33", size = 13, face = "bold")
  ) %>%
  config(displayModeBar = FALSE)
```